Constraints on Assembly Bias from Galaxy Clustering

### 2016-06-17 ms.tex
Small changes to Section 3 and acknowledgements

We constrain HOD parameters based on SDSS DR7 measurements of the projected
galaxy two-point functions, $w_{\rm p}(r_{\rm p})$, and galaxy number densities, $n_{\rm g}$,
for luminosity threshold samples published in \citet{zehavi_etal11}. We use the full covariance
matrix of the projected correlation function available at
{\tt astroweb.cwru.edu/izehavi/dr7\_covar/table8}. We assume a likelihood of the
form $\mathcal{L} \propto e^{-\chi^2/2}$, where
%
\begin{equation}
\label{eq:chisquare}
\chi^2 = \sum_{i,j} \Delta w_{{\rm p},i} \, [C^{-1}]_{ij} \, \Delta w_{{\rm p},j} + \frac{(n_{\rm g}^{\rm mock} - n_{\rm g}^{\rm meas})^2}{\sigma_{\rm n}^2},
\end{equation}
%
$\Delta w_{{\rm p},i} = w_{\rm p}^{\rm mock}(r_{{\rm p},i}) - w_{\rm p}^{\rm meas}(r_{{\rm p},i})$ is the difference
between the projected two-point function predicted by the mock catalog and the measured value in the i$^{\rm th}$
separation bin (of 12 bins total), $C^{-1}$ is the inverse of the covariance matrix of the measurements,
and the last term in Eq.~(\ref{eq:chisquare}) is the contribution from the difference between the predicted and measured
galaxy number density. The error on the galaxy number density assumes Poisson statistics for both the measured and predicted galaxy number densities.

@@

Look for the purple banana\footnote{'Til they put us in the truck.} in the third panel from the left in 
the bottom row of Figure~\ref{fig:Mr20triangle}. This panel illustrates that the parameters
$\log (M_{\rm min})$ and $\sigma_{\log M}$ share a relatively narrow degeneracy as well.
This degeneracy is largely induced by the measured
number density of the sample. Increasing $\log (M_{\rm min})$ decreases galaxy
number density, but this can be compensated by an increase in $\sigma_{\log M}$, which
places galaxies in a fraction of the considerably more numerous halos with masses less
than $M_{\rm min}$. The consequence is that $\log (M_{\rm min})$ and $\sigma_{\log M}$ are
degenerate with each other such that most of the posterior probability lies in a narrow band
along which $\log (M_{\rm min})$ and $\sigma_{\log M}$ are positively correlated.
In the following plots, we exclude the parameter $\sigma_{\log M}$,
in order to increase the clarity of the plots, because the viable range of $\sigma_{\log M}$ is
determined by this simple degeneracy with $\log (M_{\rm min})$.

The results of this subsection demonstrate that we achieve reasonable fits to projected galaxy clustering
data using direct HOD population of a high-resolution numerical simulation of structure formation. These results
also update and supersede existing constraints in the literature in at least three respects. First, we work within the
best-fit Planck cosmology. Second, we perform our parameter inference analysis using direct population of halos
identified in a numerical simulation of cosmological structure formation (BolshoiP). This greatly mitigates modeling
uncertainties associated with nonlinear density field evolution, scale-dependent halo bias, halo exclusion,
or other effects that have been difficult to incorporate into analytical halo models with high precision. Third,
we have explored the posteriors of the parameters with significantly more samples (roughly two orders of magnitude),
thereby mitigating errors on inferred parameters caused by insufficient sampling of the
posterior by \citet{zehavi_etal11}.

@@

We turn now to a discussion of our parameter inference analysis of projected galaxy clustering
in decorated HOD models that include a treatment of galaxy assembly bias. In this work,
we consider only the simplest model of galaxy assembly bias, introducing only two new
parameters, $A_{\rm cen}$ and $A_{\rm sat}$, that describe the strength of central galaxy
and satellite galaxy assembly bias respectively. These parameters are limited to values
of $-1 \le A_{\rm cen,sat} \le 1$, and $A_{\rm cen,sat}=0$ when there is no galaxy assembly bias.
In this work, we use halo concentration as our secondary halo property, so $A_{\rm cen,sat}=1$
($A_{\rm cen,sat}=-1$) means that the mean number of galaxies per halo is maximally
correlated (anti-correlated) with halo concentration. The model and its implementation
in {\tt Halotools} \cite{halotools} is discussed further in Section~\ref{subsection:decorated} above 
and in more detail in \citet{hearin_etal16}.

Examples of our fits are given in the right-hand panels of Figure~\ref{fig:Mr19samples},
Figure~\ref{fig:Mr20samples}, and Figure~\ref{fig:Mr21samples}. The general trend that
can be gleaned from these figures is that introducing assembly bias improves the ability
of the predicted two-point functions to match the measured two-point functions across the
transition from the one-halo (highly nonlinear) to two-halo (nearly linear) regimes near 
$\rp \sim 2\, h^{-1}{\mathrm{Mpc}}$, an effect that could be anticipated by the known 
scale-dependence of galaxy assembly bias \citep{sunayama_etal16}. 
This is most apparent for the $M_r < -20$ threshold sample shown in Fig.~\ref{fig:Mr20samples}.
Visually, these differences appear to be small; however, Table~\ref{table:parameters} shows that
they are statistically important.

@@

Focusing attention on the parameters describing galaxy assembly bias, it is
evident that these parameters are often quite poorly constrained by galaxy
clustering data. This is important as it implies that galaxy clustering of the precision
of SDSS DR7 measurements cannot rule out, or strongly restrict galaxy assembly
bias in many cases. This is the case, for example, for the $M_r<-21$ sample with posteriors
shown in Fig.~\ref{fig:Mr21ABtriangle}. In this case, both $A_{\rm cen}$ and
$A_{\rm sat}$ have posteriors that peak near zero, but are very broad.
Nonetheless, it is apparent that the presence of galaxy assembly bias
can alter the inferred HOD, or more generally, the inferred relationship between
galaxies and halo mass. This is most evident for the $M_r < -20$ threshold sample,
for which there are significant differences in the inferred values of all baseline HOD
parameters between models with and without galaxy assembly bias. Other threshold
samples exhibit significant differences particularly for $\alpha$, and to a lesser degree
for $\log (M_{\rm min})$ and $\sigma_{\log M}$.

@@

We thank Zheng Zheng for his cooperation in diagnosing the
differences between the present analysis and the SDSS DR7
analysis presented in \citet{zehavi_etal11}. We thank Manodeep 
Sinha for help installing and running the {\tt CorrFunc} package 
as well as comments on an earlier draft of this manuscript.
The authors gratefully acknowledge the Gauss Centre for Supercomputing
e.V. (www.gauss-centre.eu) and the Partnership for Advanced
Supercomputing in Europe (PRACE, www.prace-ri.eu) for funding the
MultiDark simulation project by providing computing time on the GCS
Supercomputer SuperMUC at Leibniz Supercomputing Centre (LRZ,
www.lrz.de). The Bolshoi simulations have been performed within the
Bolshoi project of the University of California High-Performance
AstroComputing Center (UC-HiPACC) and were run at the NASA Ames
Research Center. The work of ARZ was supported by the US National Science
Foundation via grant AST 1517563. FvdB is supported by the US National Science
Foundation through grant AST 1516962. APH is funded through the Yale Center 
for Astronomy \& Astrophysics Prize fellowship.

>>>>

We constrain HOD parameters based on SDSS DR7 measurements of the projected
galaxy two-point functions, $w_{\rm p}(r_{\rm p})$, and galaxy number densities, $n_{\rm g}$,
for luminosity threshold samples published in \citet{zehavi_etal11}. We use the full covariance
matrix of the projected correlation function available at
{\tt astroweb.cwru.edu/izehavi/dr7\_covar/table8}. We assume a likelihood of the
form $\mathcal{L} \propto e^{-\chi^2/2}$, where
%
\begin{equation}
\label{eq:chisquare}
\chi^2 = \Delta w_{{\rm p},i} \, [C^{-1}]_{ij} \, \Delta w_{{\rm p},j} + \frac{(n_{\rm g}^{\rm mock} - n_{\rm g}^{\rm meas})^2}{\sigma_{\rm n}^2},
\end{equation}
%
$\Delta w_{{\rm p},i} = w_{\rm p}^{\rm mock}(r_{{\rm p},i}) - w_{\rm p}^{\rm meas}(r_{{\rm p},i})$ is the difference
between the projected two-point function predicted by the mock catalog and the measured value in the i$^{\rm th}$
separation bin (of 12 bins total), $C^{-1}$ is the inverse of the covariance matrix of the measurements,
and the last term in Eq.~(\ref{eq:chisquare}) is the contribution from the difference between the predicted and measured
galaxy number density. In Eq.~(\ref{eq:chisquare}), the we adopt the Einstein summation convention and sum over the 
dummy indices $i$ and $j$. The error on the galaxy number density assumes Poisson statistics for both the measured and predicted galaxy number densities.

@@

Look for the purple banana\footnote{'Til they put us in the truck.} in the third panel from the left in 
the bottom row of Figure~\ref{fig:Mr20triangle}. This panel illustrates that the parameters
$\log (M_{\rm min})$ and $\sigma_{\log M}$ share a relatively narrow degeneracy as well.
This degeneracy is largely induced by the measured
number density of the sample. Increasing $\log (M_{\rm min})$ decreases galaxy
number density, but this can be compensated by an increase in $\sigma_{\log M}$, which
places galaxies in a fraction of the considerably more numerous halos with masses less
than $M_{\rm min}$. The consequence is that $\log (M_{\rm min})$ and $\sigma_{\log M}$ are
degenerate with each other such that most of the posterior probability lies in a narrow band
along which $\log (M_{\rm min})$ and $\sigma_{\log M}$ are positively correlated.
In the following plots, we exclude the parameter $\sigma_{\log M}$,
in order to increase the clarity of the plots, because the viable range of $\sigma_{\log M}$ is
determined by this simple degeneracy with $\log (M_{\rm min})$. Our constraints on $\sigma_{\log M}$ are listed in Table~\ref{tab:parameters}.

The results of this subsection demonstrate that we achieve reasonable fits to projected galaxy clustering
data using direct HOD population of a high-resolution numerical simulation of structure formation. These results
also update and supersede existing constraints in the literature in at least three respects. First, we work within the
best-fit Planck cosmology. Second, we perform our parameter inference analysis using direct population of halos
identified in a numerical simulation of cosmological structure formation (BolshoiP). This greatly mitigates modeling
uncertainties associated with nonlinear density field evolution, scale-dependent halo bias, halo exclusion,
or other effects that have been difficult to incorporate into analytical halo models with high precision
\citep[e.g.,][and references therein]{vdBosch13}. Third,
we have explored the posteriors of the parameters with significantly more samples (roughly two orders of magnitude),
thereby mitigating errors on inferred parameters caused by insufficient sampling of the
posterior by \citet{zehavi_etal11}.

@@

We turn now to a discussion of our parameter inference analysis of projected galaxy clustering
in decorated HOD models that include a treatment of galaxy assembly bias. In this work,
we consider only the simplest model of galaxy assembly bias, introducing only two new
parameters, $A_{\rm cen}$ and $A_{\rm sat}$, that describe the strength of central galaxy
and satellite galaxy assembly bias respectively. These parameters are limited to values
of $-1 \le A_{\rm cen,sat} \le 1$, and $A_{\rm cen,sat}=0$ when there is no galaxy assembly bias.
In this work, we use halo concentration as our secondary halo property, so $A_{\rm cen,sat}=1$
($A_{\rm cen,sat}=-1$) means that the mean number of galaxies per halo is maximally
correlated (anti-correlated) with halo concentration. The model and its implementation
in {\tt Halotools} is discussed further in Section~\ref{subsection:decorated} above 
and in more detail in \citet{hearin_etal16}.

Examples of our fits are given in the right-hand panels of Figure~\ref{fig:Mr19samples},
Figure~\ref{fig:Mr20samples}, and Figure~\ref{fig:Mr21samples}. The general trend that
can be gleaned from these figures is that introducing assembly bias improves the fit to
the measured two-point functions across the
transition from the one-halo (highly nonlinear) to two-halo (nearly linear) regimes near 
$\rp \sim 2\, h^{-1}{\mathrm{Mpc}}$, an effect that could be anticipated by the known 
scale-dependence of galaxy assembly bias \citep{sunayama_etal16}. 
This is most apparent for the $M_r < -20$ threshold sample shown in Fig.~\ref{fig:Mr20samples}.
Visually, these differences appear to be small; however, Table~\ref{table:parameters} shows that
they are statistically important.

@@

Focusing attention on the parameters describing galaxy assembly bias, it is
evident that these parameters are often quite poorly constrained by galaxy
clustering data. This is important as it implies that galaxy clustering of the precision
of SDSS DR7 measurements cannot rule out, or strongly restrict, galaxy assembly
bias in many cases. This is the case, for example, for the $M_r<-21$ sample with posteriors
shown in Fig.~\ref{fig:Mr21ABtriangle}. In this case, both $A_{\rm cen}$ and
$A_{\rm sat}$ have posteriors that peak near zero, but are very broad.
Nonetheless, it is apparent that the presence of galaxy assembly bias
can alter the inferred HOD, or more generally, the inferred relationship between
galaxies and halo mass. This is most evident for the $M_r < -20$ threshold sample,
for which there are significant differences in the inferred values of all baseline HOD
parameters between models with and without galaxy assembly bias. Other threshold
samples exhibit significant differences particularly for $\alpha$, and to a lesser degree
for $\log (M_{\rm min})$ and $\sigma_{\log M}$.

@@

We thank Zheng Zheng for his help in diagnosing the
differences between the present analysis and the SDSS DR7
analysis presented in \citet{zehavi_etal11}. We thank Manodeep 
Sinha for help installing and running the {\tt CorrFunc} package 
as well as comments on an earlier draft of this manuscript.
The authors gratefully acknowledge the Gauss Centre for Supercomputing
e.V. (www.gauss-centre.eu) and the Partnership for Advanced
Supercomputing in Europe (PRACE, www.prace-ri.eu) for funding the
MultiDark simulation project by providing computing time on the GCS
Supercomputer SuperMUC at Leibniz Supercomputing Centre (LRZ,
www.lrz.de). The Bolshoi simulations have been performed within the
Bolshoi project of the University of California High-Performance
AstroComputing Center (UC-HiPACC) and were run at the NASA Ames
Research Center. The work of ARZ was supported by the US National Science
Foundation via grant AST 1517563. FvdB is supported by the US National Science
Foundation through grant AST 1516962. APH is funded through the Yale Center 
for Astronomy \& Astrophysics Prize fellowship.
